---
title: "Appendix B — R Companion: Probability Concepts for Econometrics"
subtitle: "Additional student resource for Lecture 02"
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: true
    code-overflow: wrap
    smooth-scroll: true
execute:
  echo: true
  warning: false
  message: false
editor: visual
---

# How to use this companion

This is an **R companion to Appendix B: Probability Concepts** in *Principles of Econometrics, 5th edition*. It is not intended to replace the textbook and it is not a list of formulas to memorise.

The purpose is to connect probability concepts to the questions that matter later in econometrics:

- What is random in an economic model?
- What changes when we condition on information?
- What does $E(Y\mid X)$ mean?
- How can total variation be separated into systematic and remaining variation?
- What does correlation measure --- and what does it miss?
- Why do probability distributions and random-number generation matter for estimation and inference?

::: {.callout-important title="The central econometric idea"}
A regression is usually not trying to predict every individual outcome exactly. It describes systematic features of the **conditional distribution** of an outcome, especially the conditional mean $E(Y\mid X)$.
:::

## Textbook map

The companion emphasizes the following parts of Appendix B:

| Companion topic | Textbook section |
|---|---|
| Random variables, expectation, variance | B.1--B.1.2 |
| Joint, marginal and conditional distributions | B.1.3 |
| Covariance and correlation | B.1.5 |
| Conditional expectation | B.1.6 |
| Iterated expectations | B.1.7 |
| Variance decomposition | B.1.8 |
| Continuous random variables | B.2 |
| Important probability distributions | B.3 |
| Random numbers and inversion | B.4 |

Sections 1--6 are the **econometric core**. Sections 7--10 are primarily reference and exploration material.

# Random variables: concepts and realizations

A **random variable** is a variable whose value is unknown until it is observed. In economics, examples include:

- whether a consumer buys a product;
- the number of unemployment claims in a week;
- hourly wages;
- household expenditure;
- the number of visits to a shopping centre.

The important distinction is

$$
X = \text{the random variable},
\qquad
x = \text{a realized value of } X.
$$

For example, $WAGE$ can be treated as a random variable, while NOK 285 per hour is one realization.

## A simple discrete distribution

Suppose $X$ is the number of courses failed by a randomly selected student:

```{r}
#| label: discrete-example
#| fig-cap: "A simple discrete probability distribution."

x_values <- c(0, 1, 2)
probabilities <- c(0.70, 0.20, 0.10)

# A valid probability distribution must sum to one
sum(probabilities)

barplot(
  probabilities,
  names.arg = x_values,
  main = "Distribution of Failed Courses",
  xlab = "Number of courses failed",
  ylab = "Probability"
)
```

For a discrete random variable, the textbook writes

$$
f(x)=P(X=x),
$$

with

$$
0\leq f(x)\leq1,
\qquad
\sum_x f(x)=1.
$$

::: {.callout-note title="Terminology"}
The textbook uses **pdf** for discrete as well as continuous random variables. Many statistics texts call the discrete version a **probability mass function (pmf)**. The probability calculations are the same.
:::

# Expected value and variance

## Expected value

The expected value is a probability-weighted average:

$$
E(X)=\mu_X=\sum_x x f(x).
$$

```{r}
#| label: discrete-expectation

expected_value <- sum(x_values * probabilities)
expected_value
```

The result does **not** mean that a randomly selected student must fail exactly that many courses. Expectation is a property of the distribution, not a guaranteed realization.

Useful rules include

$$
E(aX+b)=aE(X)+b,
$$

and, for any two random variables,

$$
E(X+Y)=E(X)+E(Y).
$$

If $X$ and $Y$ are independent, then

$$
E(XY)=E(X)E(Y).
$$

## Variance and standard deviation

Variance measures dispersion around the expected value:

$$
\operatorname{Var}(X)
=E[(X-\mu_X)^2]
=E(X^2)-\mu_X^2.
$$

```{r}
#| label: discrete-variance

E_X2 <- sum(x_values^2 * probabilities)
variance_X <- E_X2 - expected_value^2
sd_X <- sqrt(variance_X)

c(
  expectation = expected_value,
  variance = variance_X,
  standard_deviation = sd_X
)
```

For constants $a$ and $b$,

$$
\operatorname{Var}(aX+b)=a^2\operatorname{Var}(X).
$$

For two random variables,

$$
\operatorname{Var}(X+Y)
=
\operatorname{Var}(X)
+
\operatorname{Var}(Y)
+
2\operatorname{Cov}(X,Y).
$$

::: {.callout-tip title="Why this matters for econometrics"}
Later we will distinguish between variation that a regression can account for systematically and variation that remains around the fitted relationship. Variance is therefore not just a descriptive statistic; it becomes part of the logic of regression.
:::

# Joint, marginal and conditional distributions

Probability becomes especially useful for econometrics when we move from one random variable to **two variables observed jointly**.

For discrete $X$ and $Y$,

$$
f(x,y)=P(X=x,Y=y)
$$

is the joint distribution. Marginal distributions are obtained by summing over the other variable.

## Example: employment status and gender

Consider this hypothetical population joint distribution:

```{r}
#| label: joint-distribution

employment_data <- matrix(
  c(0.30, 0.20,
    0.15, 0.35),
  nrow = 2,
  dimnames = list(
    Gender = c("Male", "Female"),
    Employment = c("Employed", "Unemployed")
  )
)

employment_data
sum(employment_data)
```

The marginal distributions are

```{r}
#| label: marginal-distributions

margin_gender <- rowSums(employment_data)
margin_employment <- colSums(employment_data)

margin_gender
margin_employment
```

## Conditioning means using information

Without information about gender, the probability of unemployment is

```{r}
#| label: conditional-probability

p_unemployed <- margin_employment["Unemployed"]
p_unemployed
```

Conditional on the individual being female,

$$
P(U\mid Female)
=
\frac{P(U,Female)}{P(Female)}.
$$

```{r}
p_unemployed_given_female <-
  employment_data["Female", "Unemployed"] /
  margin_gender["Female"]

c(
  P_unemployed = p_unemployed,
  P_unemployed_given_female = p_unemployed_given_female
)
```

The probability changes because the information set changes.

::: {.callout-important title="Econometric bridge"}
This is the same conceptual move we make when going from $E(Y)$ to $E(Y\mid X)$. **Conditioning means using information about $X$ to describe the distribution of $Y$.**
:::

## Independence

$X$ and $Y$ are independent if

$$
f(x,y)=f_X(x)f_Y(y)
$$

for every possible pair $(x,y)$.

The joint distribution that would arise under independence is

```{r}
#| label: independence-check

joint_if_independent <- outer(margin_gender, margin_employment)

joint_if_independent
employment_data
```

The two tables differ, so the variables in this hypothetical example are not independent.

# Conditional expectation: the foundation for regression

The **conditional expectation**

$$
E(Y\mid X=x)
$$

is the average value of $Y$ among observations for which $X=x$.

This is one of the most important ideas in introductory econometrics.

## A simulated income--education population

We construct an economic example where income varies systematically with education but also contains substantial individual variation.

```{r}
#| label: income-data

set.seed(131415)

n <- 5000
education_levels <- 8:20

# These probabilities sum to exactly one
education_probs_population <- c(
  0.03, 0.04, 0.05, 0.06, 0.07,
  0.09, 0.12, 0.14, 0.12, 0.10,
  0.08, 0.06, 0.04
)

sum(education_probs_population)

education <- sample(
  education_levels,
  size = n,
  replace = TRUE,
  prob = education_probs_population
)

# Systematic component plus individual variation
income <- 20000 + 3000 * education + rnorm(n, mean = 0, sd = 10000)

income_data <- data.frame(
  education = education,
  income = income
)
```

The data-generating relationship was chosen so that

$$
E(INCOME\mid EDUCATION)
=
20000+3000\,EDUCATION.
$$

But an individual person's income is not determined by education alone.

## Estimate conditional means

```{r}
#| label: conditional-means

conditional_means <- with(
  income_data,
  tapply(income, education, mean)
)

round(conditional_means, 0)
```

Visualize the full conditional distributions, not only the means:

```{r}
#| label: income-boxplot
#| fig-cap: "Income distributions conditional on education. Red points show conditional means."

boxplot(
  income ~ education,
  data = income_data,
  main = "Income Distribution by Education Level",
  xlab = "Years of education",
  ylab = "Income"
)

points(
  seq_along(conditional_means),
  conditional_means,
  pch = 19,
  cex = 1.15
)
```

::: {.callout-warning title="Do not confuse a conditional mean with an individual outcome"}
If $E(INCOME\mid EDUCATION=16)=70{,}000$, this does **not** imply that every person with 16 years of education earns 70,000. It describes the centre of a conditional distribution.
:::

# Law of iterated expectations

The law of iterated expectations states

$$
E(Y)=E_X[E(Y\mid X)].
$$

In words:

> The unconditional mean can be recovered by averaging the conditional means, using the distribution of $X$ as weights.

Using the simulated income data:

```{r}
#| label: iterated-expectations

education_probs_sample <- prop.table(table(income_data$education))

mean_from_conditionals <- sum(
  conditional_means * education_probs_sample
)

overall_mean <- mean(income_data$income)

c(
  overall_mean = overall_mean,
  mean_from_conditional_means = mean_from_conditionals
)
```

The two values are the same apart from numerical rounding because every observation is represented in exactly one education group.

::: {.callout-tip title="Interpretation"}
The law is mathematically simple but conceptually powerful: an unconditional population quantity can often be understood by first conditioning on economically relevant information and then averaging back over that information.
:::

# Variance decomposition: between and within variation

Appendix B gives the identity

$$
\operatorname{Var}(Y)
=
\operatorname{Var}_X[E(Y\mid X)]
+
E_X[\operatorname{Var}(Y\mid X)].
$$

Think of this as

$$
\boxed{
\text{Total variation}
=
\text{between-group variation}
+
\text{within-group variation}
}
$$

For the identity to hold exactly in our finite teaching dataset, we calculate **population variances** using denominator $n$, rather than R's default sample variance with denominator $n-1$.

```{r}
#| label: population-variance-helper

pop_var <- function(z) {
  mean((z - mean(z))^2)
}
```

## Exact decomposition in R

```{r}
#| label: variance-decomposition

# Conditional variances calculated as population variances within each group
conditional_vars <- with(
  income_data,
  tapply(income, education, pop_var)
)

# Realized probability of each education group in this sample
p_x <- prop.table(table(income_data$education))

# Total population variance
var_total <- pop_var(income_data$income)

# Variance of E(Y|X): weighted variation in conditional means
var_between <- sum(
  p_x * (conditional_means - mean(income_data$income))^2
)

# E[Var(Y|X)]: weighted average of within-group variances
var_within <- sum(
  p_x * conditional_vars
)

c(
  total = var_total,
  between = var_between,
  within = var_within,
  between_plus_within = var_between + var_within
)
```

The numerical identity should satisfy

$$
\operatorname{Var}(Y)
\approx
\operatorname{Var}[E(Y\mid X)]
+
E[\operatorname{Var}(Y\mid X)].
$$

The proportion of total variation occurring between education-group means is

```{r}
#| label: variance-share

between_share <- var_between / var_total
between_share
```

## Connection to regression and $R^2$

Now estimate a regression with a separate indicator for every education category:

```{r}
#| label: group-regression

model_group <- lm(income ~ factor(education), data = income_data)
r2_group <- summary(model_group)$r.squared

c(
  variance_share = between_share,
  regression_R2 = r2_group
)
```

These numbers agree because this regression reproduces the observed conditional mean for each education group. The fitted values are therefore

$$
\widehat Y_i = \widehat E(Y\mid X_i).
$$

The usual regression decomposition

$$
TSS = ESS + RSS
$$

is closely related to the probability decomposition between systematic and remaining variation.

### A specification choice: group means or a straight line?

A separate mean for every education level is flexible. A simple regression instead imposes a linear pattern:

```{r}
#| label: linear-vs-group

model_linear <- lm(income ~ education, data = income_data)

c(
  R2_group_means = summary(model_group)$r.squared,
  R2_linear = summary(model_linear)$r.squared
)
```

::: {.callout-important title="Econometric judgment"}
The probability model tells us what conditional means are. Regression adds a **modeling choice** about how those conditional means depend on $X$. A straight line is one possible specification, not a mathematical necessity.
:::

# Covariance and correlation

Covariance measures linear comovement:

$$
\operatorname{Cov}(X,Y)
=
E[(X-\mu_X)(Y-\mu_Y)]
=
E(XY)-\mu_X\mu_Y.
$$

Correlation standardizes covariance:

$$
\rho_{XY}
=
\frac{\operatorname{Cov}(X,Y)}{\sigma_X\sigma_Y},
\qquad
-1\leq\rho\leq1.
$$

Correlation measures the strength of a **linear** relationship.

## A conventional linear example

```{r}
#| label: correlation-linear
#| fig-cap: "A strong positive linear association."

set.seed(123)

x_linear <- rnorm(1000, mean = 50, sd = 10)
y_linear <- 2 * x_linear + rnorm(1000, mean = 0, sd = 20)

c(
  covariance = cov(x_linear, y_linear),
  correlation = cor(x_linear, y_linear)
)

plot(
  x_linear,
  y_linear,
  pch = 16,
  col = adjustcolor("black", alpha.f = 0.30),
  xlab = "X",
  ylab = "Y",
  main = "Linear Association"
)
abline(lm(y_linear ~ x_linear), lwd = 2)
```

## The zero-correlation trap

Before running the next example, predict what the correlation should be.

```{r}
#| label: correlation-nonlinear
#| fig-cap: "Zero correlation does not imply independence."

set.seed(321)

x_nonlinear <- runif(2000, -1, 1)
y_nonlinear <- x_nonlinear^2

cor(x_nonlinear, y_nonlinear)

plot(
  x_nonlinear,
  y_nonlinear,
  pch = 16,
  col = adjustcolor("black", alpha.f = 0.25),
  xlab = "X",
  ylab = "Y = X²",
  main = "A Strong Nonlinear Relationship"
)
```

$Y$ is completely determined by $X$, yet the correlation is approximately zero because positive and negative linear comovement cancel.

::: {.callout-warning title="Important distinction"}
Independence implies zero covariance (when the moments exist), but zero covariance or zero correlation does **not** generally imply independence.
:::

::: {.callout-note title="Also remember"}
Correlation and conditional association are descriptive. Neither one, by itself, establishes a causal effect.
:::

# Reference: important discrete distributions

The remaining sections are useful reference material. You do not need to memorise every formula before it becomes relevant in an econometric problem.

## Bernoulli distribution

A Bernoulli random variable has two outcomes:

$$
P(X=1)=p,
\qquad
P(X=0)=1-p.
$$

Its moments are

$$
E(X)=p,
\qquad
\operatorname{Var}(X)=p(1-p).
$$

Bernoulli variables are especially important for binary outcomes and indicator variables.

## Binomial distribution

If $X$ counts successes in $n$ independent Bernoulli trials with common success probability $p$,

$$
X\sim Binomial(n,p),
$$

with

$$
E(X)=np,
\qquad
\operatorname{Var}(X)=np(1-p).
$$

### R example: customer purchases

```{r}
#| label: binomial-example
#| fig-cap: "Binomial distribution for the number of purchasers among 20 customers."

n_customers <- 20
prob_purchase <- 0.30
k_values <- 0:n_customers

prob_k <- dbinom(
  k_values,
  size = n_customers,
  prob = prob_purchase
)

barplot(
  prob_k,
  names.arg = k_values,
  main = "Binomial Distribution: n = 20, p = 0.30",
  xlab = "Number of purchases",
  ylab = "Probability"
)

c(
  theoretical_mean = n_customers * prob_purchase,
  theoretical_variance = n_customers * prob_purchase * (1 - prob_purchase)
)
```

Simulate repeated groups of 20 customers:

```{r}
#| label: binomial-simulation

set.seed(456)

simulated_purchases <- rbinom(
  5000,
  size = n_customers,
  prob = prob_purchase
)

c(
  simulated_mean = mean(simulated_purchases),
  simulated_variance = var(simulated_purchases)
)
```

# Reference: continuous random variables

A continuous random variable can take values in an interval. For any particular point,

$$
P(X=c)=0.
$$

Probabilities refer to intervals and are areas under the density:

$$
P(a\leq X\leq b)=\int_a^b f(x)\,dx.
$$

A valid continuous density satisfies

$$
f(x)\geq0
$$

and

$$
\int_{-\infty}^{\infty}f(x)\,dx=1.
$$

## Example: $f(x)=2(1-x)$ on $[0,1]$

```{r}
#| label: continuous-density
#| fig-cap: "The density f(x) = 2(1-x) on [0,1]."

x_grid <- seq(0, 1, length.out = 1000)
density_grid <- 2 * (1 - x_grid)

plot(
  x_grid,
  density_grid,
  type = "l",
  lwd = 2,
  xlab = "x",
  ylab = "f(x)",
  main = "Continuous Density: f(x) = 2(1-x)"
)

integrate(
  function(x) 2 * (1 - x),
  lower = 0,
  upper = 1
)$value
```

Expectation and variance are obtained by integration:

$$
E(X)=\int x f(x)\,dx,
$$

$$
\operatorname{Var}(X)=E(X^2)-[E(X)]^2.
$$

```{r}
#| label: continuous-moments

E_X <- integrate(
  function(x) x * 2 * (1 - x),
  lower = 0,
  upper = 1
)$value

E_X2 <- integrate(
  function(x) x^2 * 2 * (1 - x),
  lower = 0,
  upper = 1
)$value

Var_X <- E_X2 - E_X^2

c(
  expectation = E_X,
  variance = Var_X
)
```

# Reference: simulation by inversion

For the density

$$
f(x)=2(1-x),\qquad0\leq x\leq1,
$$

the cumulative distribution function is

$$
F(x)=2x-x^2.
$$

If $U\sim Uniform(0,1)$ and $U=F(X)$, solving for $X$ gives

$$
X=1-\sqrt{1-U}.
$$

This is the **inversion method** discussed in Appendix B.4.

```{r}
#| label: inversion-simulation
#| fig-cap: "Simulation from f(x) = 2(1-x) using the inversion method."

set.seed(789)

n_sim <- 100000
u <- runif(n_sim)
x_sim <- 1 - sqrt(1 - u)

c(
  theoretical_mean = 1 / 3,
  simulated_mean = mean(x_sim),
  theoretical_variance = 1 / 18,
  simulated_variance = var(x_sim)
)

hist(
  x_sim,
  breaks = 40,
  probability = TRUE,
  main = "Simulation by Inversion",
  xlab = "X"
)

curve(
  2 * (1 - x),
  from = 0,
  to = 1,
  add = TRUE,
  lwd = 2
)
```

::: {.callout-tip title="Why this version is preferable to rejection sampling here"}
The inversion approach uses the CDF directly, matches Appendix B.4, and avoids a slow loop that repeatedly appends one accepted draw at a time.
:::

# Reference: uniform and normal distributions

## Uniform distribution

For $X\sim Uniform(a,b)$,

$$
f(x)=\frac{1}{b-a},
\qquad a\leq x\leq b,
$$

$$
E(X)=\frac{a+b}{2},
\qquad
\operatorname{Var}(X)=\frac{(b-a)^2}{12}.
$$

```{r}
#| label: uniform-simulation

set.seed(2026)

uniform_samples <- runif(10000)

c(
  theoretical_mean = 0.5,
  empirical_mean = mean(uniform_samples),
  theoretical_variance = 1 / 12,
  empirical_variance = var(uniform_samples)
)
```

## Normal distribution

If

$$
X\sim N(\mu,\sigma^2),
$$

the distribution is symmetric around $\mu$, with

$$
E(X)=\mu,
\qquad
\operatorname{Var}(X)=\sigma^2.
$$

```{r}
#| label: normal-comparison
#| fig-cap: "Two normal distributions with different means and variances."

x_range <- seq(-4, 8, length.out = 1000)

y_standard <- dnorm(x_range, mean = 0, sd = 1)
y_shifted <- dnorm(x_range, mean = 2, sd = 1.5)

plot(
  x_range,
  y_standard,
  type = "l",
  lwd = 2,
  xlab = "x",
  ylab = "Density",
  main = "Comparison of Normal Distributions"
)

lines(x_range, y_shifted, lwd = 2, lty = 2)

legend(
  "topright",
  legend = c("N(0,1)", "N(2,2.25)"),
  lwd = 2,
  lty = c(1, 2)
)
```

R provides four useful families of functions for standard distributions:

- `d...` for densities or probability masses;
- `p...` for cumulative probabilities;
- `q...` for quantiles;
- `r...` for random draws.

For the normal distribution these are `dnorm()`, `pnorm()`, `qnorm()` and `rnorm()`.

```{r}
#| label: normal-functions

mean_score <- 75
sd_score <- 10

prob_above_85 <- 1 - pnorm(
  85,
  mean = mean_score,
  sd = sd_score
)

score_90th <- qnorm(
  0.90,
  mean = mean_score,
  sd = sd_score
)

c(
  P_score_above_85 = prob_above_85,
  percentile_90 = score_90th
)
```

::: {.callout-warning title="Normality and regression"}
Do not conclude that OLS requires the regression errors themselves to be normally distributed. Normality becomes useful for some exact finite-sample inference results, but it is **not required to calculate OLS estimates**.
:::

# Reference: pseudo-random numbers

Computers use deterministic algorithms, so computer-generated random numbers are **pseudo-random**. For empirical work, the main practical lessons are:

1. use established random-number routines;
2. use `set.seed()` when reproducibility matters;
3. document software and code used for simulation.

## A simple linear congruential generator

Appendix B describes recursive generators of the form

$$
X_t=(aX_{t-1}+c)\bmod m.
$$

The following code is for illustration only; use R's established generators in actual work.

```{r}
#| label: simple-rng

simple_rng <- function(seed, n,
                       a = 1664525,
                       c = 1013904223,
                       m = 2^32) {
  x <- numeric(n)
  x[1] <- seed

  for (i in 2:n) {
    x[i] <- (a * x[i - 1] + c) %% m
  }

  x / m
}

custom_random <- simple_rng(seed = 12345, n = 10000)

c(
  mean = mean(custom_random),
  variance = var(custom_random),
  minimum = min(custom_random),
  maximum = max(custom_random)
)
```

## Inversion with another distribution

For an exponential distribution with rate $\lambda$,

$$
F^{-1}(u)=\frac{-\log(1-u)}{\lambda}.
$$

```{r}
#| label: exponential-inversion
#| fig-cap: "Exponential random numbers generated by inversion and by R's built-in generator."

set.seed(20260811)

lambda <- 2
uniform_vals <- runif(5000)
exponential_inversion <- -log(1 - uniform_vals) / lambda
exponential_builtin <- rexp(5000, rate = lambda)

par(mfrow = c(1, 2))

hist(
  exponential_inversion,
  breaks = 35,
  probability = TRUE,
  main = "Via inversion",
  xlab = "Value"
)

hist(
  exponential_builtin,
  breaks = 35,
  probability = TRUE,
  main = "R built-in",
  xlab = "Value"
)

par(mfrow = c(1, 1))
```

# Self-check: five questions

Use these questions to check whether you have understood the probability concepts rather than only the R commands.

1. If $E(WAGE\mid EDUC=16)=300$, does a randomly selected person with 16 years of education have to earn 300? Explain.
2. What is the difference between $P(U)$ and $P(U\mid Female)$ in the employment example?
3. Why must the conditional means be **weighted** when using the law of iterated expectations?
4. In the variance decomposition, what economic interpretation can you give to the between-group and within-group components?
5. How can $Y=X^2$ have approximately zero correlation between $X$ and $Y$ even though $Y$ is completely determined by $X$?

<details>
<summary><strong>Check your answers</strong></summary>

1. No. The conditional expectation is the average or centre of the conditional distribution; individual realizations can differ substantially.
2. $P(U)$ uses no gender information. $P(U\mid Female)$ restricts the reference population to females; conditioning changes the information set.
3. Education levels occur with different probabilities. An unweighted average would incorrectly treat rare and common education groups as equally prevalent.
4. Between-group variation measures how conditional means differ across education groups; within-group variation measures the remaining dispersion among observations with the same education.
5. Correlation measures linear association. For a symmetric $X$, the positive and negative linear comovement in the U-shaped relationship can cancel even though the nonlinear dependence is exact.

</details>

# Key takeaways for econometrics

1. **Economic variables are random variables.** Observed data are realizations from distributions.
2. **Conditioning means using information.** $E(Y)$ and $E(Y\mid X)$ answer different questions.
3. **Conditional expectation is central to regression.** Regression models how the mean of $Y$ changes with information in $X$.
4. **Variation has structure.** Total variation can be decomposed into variation between conditional means and variation remaining within conditional distributions.
5. **Correlation measures linear association, not all dependence.** Zero correlation does not imply independence.
6. **Association is not causation.** Probability and regression relationships need additional assumptions before receiving a causal interpretation.
7. **Probability distributions are tools, not a catalogue to memorise.** Return to Bernoulli, normal, $t$, $\chi^2$, $F$, and related distributions when an econometric problem requires them.
8. **Simulation should be reproducible.** Use established random-number routines and `set.seed()` when appropriate.

::: {.callout-important title="The idea to carry into the next econometrics lecture"}
$$
\boxed{E(Y\mid X=x)\neq Y_i}
$$

Econometrics searches for systematic relationships in outcomes that remain uncertain.
:::
