---
title: "Seminar: Confidence Intervals and Hypothesis Testing"
subtitle: "Chapter 3"
lang: en
format:
  html:
    toc: true
    code-copy: true
    embed-resources: true
execute:
  warning: true
  message: false
---

## Learning objectives

After the seminar, you should be able to

- construct and interpret a confidence interval for a regression coefficient,
- formulate a null and an alternative hypothesis,
- carry out a one-sided or two-sided $t$-test,
- calculate and interpret a p-value, and
- state the conclusion in its economic context.

The general workflow is

$$
\text{economic question}
\longrightarrow H_0\text{ and }H_1
\longrightarrow t\text{-value}
\longrightarrow \text{p-value}
\longrightarrow \text{conclusion}.
$$
```{r}
#| label: setup
#| include: false

options(scipen = 10)
options(digits = 4)


```




## Five-minute review

### Confidence intervals

Let $b_k$ estimate $\beta_k$, with estimated standard error
$\operatorname{se}(b_k)$. A $100(1-\alpha)$ percent confidence interval is

$$
b_k\pm t(1-\alpha/2,N-2)\operatorname{se}(b_k).
$$

The method produces intervals that contain the true parameter in
$100(1-\alpha)$ percent of repeated samples. It is not correct to say that
there is a 95 percent probability that a fixed parameter lies in one already
calculated interval.

### Hypothesis tests

For $H_0:\beta_k=c$, the test statistic is

$$
t=\frac{b_k-c}{\operatorname{se}(b_k)}.
$$

If $H_0$ is true and the model assumptions hold, it follows a $t$-distribution
with $N-2$ degrees of freedom.

| Alternative | Test | Extreme outcomes are in |
|---|---|---|
| $H_1:\beta_k>c$ | Right-tailed | Right tail |
| $H_1:\beta_k<c$ | Left-tailed | Left tail |
| $H_1:\beta_k\neq c$ | Two-sided | Both tails |

The direction must be determined by the economic question before examining
the estimate.

### P-values

The p-value is the probability, **assuming that the null hypothesis is true**,
of obtaining a test statistic at least as extreme as the observed value in the
direction specified by $H_1$.

Reject $H_0$ if $p\leq\alpha$. If we do not reject $H_0$, the data do not
provide sufficient evidence against it. This does not prove that $H_0$ is
true.

## Warm-up: critical values in R

Suppose that $N=32$. In a simple linear regression with an intercept and one
explanatory variable, the degrees of freedom are

$$
df=N-2=30.
$$

Use R to find

1. the critical value for a 95 percent confidence interval,
2. the critical value for a right-tailed test with $\alpha=0.05$,
3. the critical value for a left-tailed test with $\alpha=0.05$, and
4. both critical values for a two-sided test with $\alpha=0.05$.

```{r}
#| label: critical-values-warm-up

N <- 32
df <- N - 2
alpha <- 0.05

critical_values <- c(
  confidence_interval_95 = qt(1 - alpha / 2, df = df),
  right_tailed_test = qt(1 - alpha, df = df),
  left_tailed_test = qt(alpha, df = df),
  two_sided_lower = qt(alpha / 2, df = df),
  two_sided_upper = qt(1 - alpha / 2, df = df)
)

critical_values
```

The location of the alternative hypothesis determines the tail or tails and,
therefore, which percentile is required:

$$
H_1
\longrightarrow
\text{tail or tails}
\longrightarrow
\text{critical value}.
$$

## Worked example: food expenditure and income

Consider

$$
FOOD\_EXP_i=\beta_1+\beta_2INCOME_i+e_i.
$$

`food_exp` is weekly food expenditure in dollars. `income` is weekly household
income measured in hundreds of dollars. Thus, $\beta_2$ measures the change in
expected weekly food expenditure associated with a 100-dollar increase in
weekly income.

### Estimate the model

```{r}
#| label: load-and-estimate-food

load(url("http://www.principlesofeconometrics.com/poe5/data/rdata/food.rdata"))

food_model <- lm(food_exp ~ income, data = food)
summary(food_model)

b2_food <- coef(food_model)["income"]
se_b2_food <- sqrt(vcov(food_model)["income", "income"])
df_food <- df.residual(food_model)

c(
  estimate = b2_food,
  standard_error = se_b2_food,
  degrees_of_freedom = df_food
)
```

### A 95 percent confidence interval

#### 1. Calculate it in R

```{r}
#| label: food-confidence-interval-en

alpha_ci <- 0.05
tc_ci <- qt(1 - alpha_ci / 2, df = df_food)

food_ci <- c(
  lower = b2_food - tc_ci * se_b2_food,
  estimate = b2_food,
  upper = b2_food + tc_ci * se_b2_food
)

food_ci
confint(food_model, "income", level = 0.95)
```

#### 2. Connect the code to the formula

$$
\begin{aligned}
b_2\pm t_c\operatorname{se}(b_2)
&=`r round(b2_food, 4)`
\pm `r round(tc_ci, 4)`\times`r round(se_b2_food, 4)`\\
&=[`r round(food_ci[1], 4)`,\;
`r round(food_ci[3], 4)`].
\end{aligned}
$$

#### 3. Interpret the result

A 100-dollar increase in weekly income is associated with an estimated
`r round(b2_food, 2)`-dollar increase in expected weekly food expenditure. The
95 percent confidence interval ranges from approximately
`r round(food_ci[1], 2)` to `r round(food_ci[3], 2)` dollars. This
is an estimated association; the regression alone does not establish a causal
effect.

### Test 1: Is the slope different from zero?

We test

$$
H_0:\beta_2=0
\qquad\text{against}\qquad
H_1:\beta_2\neq0.
$$

This is a two-sided test because both positive and negative departures from
zero count as evidence against $H_0$.

```{r}
#| label: food-two-sided-test-en

null_1 <- 0
alpha_1 <- 0.05

t_food_1 <- (b2_food - null_1) / se_b2_food
tc_food_1 <- qt(1 - alpha_1 / 2, df = df_food)
p_food_1 <- 2 * pt(-abs(t_food_1), df = df_food)

c(
  t_statistic = t_food_1,
  lower_critical_value = -tc_food_1,
  upper_critical_value = tc_food_1,
  p_value = p_food_1
)
```

The calculation is

$$
t=\frac{b_2-0}{\operatorname{se}(b_2)}
=\frac{`r round(b2_food, 4)`-0}{`r round(se_b2_food, 4)`}
=`r round(t_food_1, 4)`.
$$

The two-sided p-value is

$$
p=2P\left[T(38)\geq |`r round(t_food_1, 4)`|\right]
=`r format(p_food_1, scientific = TRUE, digits = 3)`.
$$

We reject $H_0$ at the 5 percent level. The data provide statistical evidence
of an association between income and expected food expenditure.

### Test 2: Is the increase greater than 5.50 dollars?

Suppose profitability requires expected food expenditure to rise by more than
5.50 dollars when weekly income rises by 100 dollars. We test

$$
H_0:\beta_2\leq5.5
\qquad\text{against}\qquad
H_1:\beta_2>5.5.
$$

```{r}
#| label: food-right-tailed-test-en

null_2 <- 5.5
alpha_2 <- 0.05

t_food_2 <- (b2_food - null_2) / se_b2_food
tc_food_2 <- qt(1 - alpha_2, df = df_food)
p_food_2 <- pt(t_food_2, df = df_food, lower.tail = FALSE)

c(
  t_statistic = t_food_2,
  critical_value = tc_food_2,
  p_value = p_food_2,
  reject_null = p_food_2 <= alpha_2
)
```

The calculation is

$$
t=\frac{b_2-5.5}{\operatorname{se}(b_2)}
=\frac{`r round(b2_food, 4)`-5.5}{`r round(se_b2_food, 4)`}
=`r round(t_food_2, 4)`.
$$

Because the test is right-tailed,

$$
p=P\left[T(38)\geq `r round(t_food_2, 4)`\right]
=`r round(p_food_2, 4)`.
$$

We reject $H_0$ at the 5 percent level, but not at the 1 percent level. The
data provide evidence at the 5 percent level that the increase exceeds 5.50
dollars.

For comparison, the two-sided p-value for the same null value is

```{r}
#| label: compare-food-p-values-en

p_food_2_two_sided <- 2 * pt(-abs(t_food_2), df = df_food)

c(
  right_tailed_p_value = p_food_2,
  two_sided_p_value = p_food_2_two_sided
)
```

The test statistic does not change. The p-value changes because the two-sided
test includes extreme outcomes in both tails.

::: {.callout-important}
## Before starting the student exercise

For every test, write down $H_0$ and $H_1$ before calculating the p-value.
Finish with one statistical conclusion and one interpretation in the economic
context.
:::

## Student exercise: transport expenditure and income

Consider

$$
TRANSPORT_i=\beta_1+\beta_2INCOME_i+e_i,
$$

where transport expenditure is measured in dollars and income in hundreds of
dollars. For $N=32$ households, the estimated regression is

$$
\widehat{TRANSPORT}=35.2+6.8INCOME,
$$

with $\operatorname{se}(b_1)=18.4$ and
$\operatorname{se}(b_2)=1.7$.

### (a) Interpretation and degrees of freedom

```{r}
#| label: transport-setup-worked

b1 <- 35.2
b2 <- 6.8
se_b1 <- 18.4
se_b2 <- 1.7
N <- 32
df <- N - 2

c(b1 = b1, b2 = b2, df = df)
```

The intercept estimates expected weekly transport expenditure to be 35.20
dollars when weekly income is zero. This interpretation should be treated
cautiously if the sample contains no households with income near zero.

Because income is measured in hundreds of dollars, the slope says that a
100-dollar increase in weekly income is associated with an estimated
6.80-dollar increase in expected weekly transport expenditure. The regression
has $32-2=30$ degrees of freedom.

### (b) A 95 percent confidence interval

```{r}
#| label: transport-confidence-interval-worked

alpha <- 0.05
tc <- qt(1 - alpha / 2, df = df)

transport_ci <- c(
  lower = b2 - tc * se_b2,
  estimate = b2,
  upper = b2 + tc * se_b2
)

transport_ci
```

Substituting the numbers into the formula gives

$$
\begin{aligned}
b_2\pm t(0.975,30)\operatorname{se}(b_2)
&=6.8\pm `r round(tc, 4)`\times1.7\\
&=[`r round(transport_ci["lower"], 4)`,\;
`r round(transport_ci["upper"], 4)`].
\end{aligned}
$$

We estimate with 95 percent confidence that a 100-dollar increase in weekly
income is associated with an increase of between approximately
`r round(transport_ci["lower"], 2)` and
`r round(transport_ci["upper"], 2)` dollars in expected weekly transport
expenditure.

### (c) A right-tailed economic hypothesis

The company's profitability claim gives

$$
H_0:\beta_2\leq4
\qquad\text{against}\qquad
H_1:\beta_2>4.
$$

```{r}
#| label: transport-right-test-worked

beta_null_right <- 4
alpha_right <- 0.05

t_transport_right <-
  (b2 - beta_null_right) / se_b2

tc_transport_right <-
  qt(1 - alpha_right, df = df)

p_transport_right <-
  pt(t_transport_right, df = df, lower.tail = FALSE)

c(
  t_statistic = t_transport_right,
  critical_value = tc_transport_right,
  p_value = p_transport_right,
  reject_null = p_transport_right <= alpha_right
)
```

The test statistic is

$$
t=\frac{6.8-4}{1.7}=`r round(t_transport_right, 4)`.
$$

The 5 percent critical value is
$t(0.95,30)=`r round(tc_transport_right, 4)`$, and the p-value is

$$
P[T(30)\geq `r round(t_transport_right, 4)`]
=`r round(p_transport_right, 4)`.
$$

Because the p-value is slightly above 0.05, we do not reject $H_0$ at the
5 percent level. The data do not provide sufficient evidence at that level
that the increase exceeds 4 dollars. Because the p-value is below 0.10, the
null hypothesis would be rejected at the 10 percent level.

### (d) A two-sided hypothesis

We test

$$
H_0:\beta_2=8
\qquad\text{against}\qquad
H_1:\beta_2\neq8.
$$

```{r}
#| label: transport-two-sided-test-worked

beta_null_two <- 8
alpha_two <- 0.05

t_transport_two <-
  (b2 - beta_null_two) / se_b2

tc_transport_two <-
  qt(1 - alpha_two / 2, df = df)

p_transport_two <-
  2 * pt(-abs(t_transport_two), df = df)

c(
  t_statistic = t_transport_two,
  lower_critical_value = -tc_transport_two,
  upper_critical_value = tc_transport_two,
  p_value = p_transport_two,
  reject_null = p_transport_two <= alpha_two
)
```

$$
t=\frac{6.8-8}{1.7}=`r round(t_transport_two, 4)`.
$$

The critical values are
$\pm t(0.975,30)=\pm`r round(tc_transport_two, 4)`$. The p-value is

$$
2P[T(30)\geq |`r round(t_transport_two, 4)`|]
=`r round(p_transport_two, 4)`.
$$

We do not reject $H_0$. The data are compatible with a population slope of 8,
but this result does not prove that the slope equals 8.

### (e) Compare the tests

The tests have different p-values because they use different null values and
different alternatives. Part (c) places all relevant probability in the right
tail, whereas part (d) uses both tails.

A result may be rejected at 10 percent but not at 5 percent because the first
rule uses a less demanding threshold. Here the right-tailed p-value lies
between 0.05 and 0.10.

For the slope, `summary(model)` would automatically report the two-sided test

$$
H_0:\beta_2=0
\qquad\text{against}\qquad
H_1:\beta_2\neq0.
$$

That is not the hypothesis used in either part (c) or part (d), so its p-value
would not answer those economic questions.

## Exit exercise: a left-tailed test

Suppose $b_2=-0.90$, $\operatorname{se}(b_2)=0.35$, and $N=25$. We test

$$
H_0:\beta_2\geq0
\qquad\text{against}\qquad
H_1:\beta_2<0.
$$

```{r}
#| label: left-tailed-exit-worked

b2_demand <- -0.90
se_b2_demand <- 0.35
df_demand <- 25 - 2
alpha_demand <- 0.05

t_demand <- b2_demand / se_b2_demand
tc_demand <- qt(alpha_demand, df = df_demand)
p_demand <- pt(t_demand, df = df_demand)

c(
  t_statistic = t_demand,
  critical_value = tc_demand,
  p_value = p_demand,
  reject_null = p_demand <= alpha_demand
)
```

The test statistic is

$$
t=\frac{-0.90-0}{0.35}=`r round(t_demand, 4)`.
$$

The left-tail critical value is
$t(0.05,23)=`r round(tc_demand, 4)`$, and the p-value is
`r round(p_demand, 4)`. We reject $H_0$ at the 5 percent level. The data
provide evidence of an inverse relationship between price and expected
demand.

## Extra Exercise 3.22: house prices in Collegetown

The dataset contains 500 sales of single-family houses in Baton Rouge between
2009 and 2013. `price` is measured in thousands of dollars, while `sqft` is
floor area measured in hundreds of square feet. We estimate

$$
PRICE_i=\beta_1+\beta_2SQFT_i+e_i.
$$

### Preparation

```{r}
#| label: load-collegetown-worked-en

load(url(
  "http://www.principlesofeconometrics.com/poe5/data/rdata/collegetown.rdata"
))

plot(
  x = collegetown$sqft,
  y = collegetown$price,
  xlab = "Floor area (hundreds of square feet)",
  ylab = "Sale price (thousands of dollars)",
  main = "House price and floor area",
  pch = 16,
  col = rgb(0, 0, 0, 0.35)
)
```

### 3.22(a): Estimate the model and elasticity

#### 1. Extract the required quantities in R

```{r}
#| label: solution-322a-en

house_model <- lm(price ~ sqft, data = collegetown)
summary(house_model)

b2_house <- coef(house_model)["sqft"]
se_b2_house <- sqrt(vcov(house_model)["sqft", "sqft"])
df_house <- df.residual(house_model)

mean_sqft <- mean(collegetown$sqft)
mean_price <- mean(collegetown$price)
elasticity_scale <- mean_sqft / mean_price

elasticity_hat <- elasticity_scale * b2_house
se_elasticity <- abs(elasticity_scale) * se_b2_house

tc_elasticity <- qt(0.975, df = df_house)
elasticity_ci <- c(
  lower = elasticity_hat - tc_elasticity * se_elasticity,
  estimate = elasticity_hat,
  upper = elasticity_hat + tc_elasticity * se_elasticity
)

c(
  mean_sqft = mean_sqft,
  mean_price = mean_price,
  slope = b2_house,
  slope_standard_error = se_b2_house
)

elasticity_ci
```

#### 2. Substitute the numbers into the formulas

The slope and standard error are
$b_2=`r round(b2_house, 4)`$ and
$\operatorname{se}(b_2)=`r round(se_b2_house, 4)`$. Therefore,

$$
\widehat\varepsilon
=`r round(b2_house, 4)`
\left(\frac{`r round(mean_sqft, 4)`}{`r round(mean_price, 4)`}\right)
=`r round(elasticity_hat, 4)`.
$$

The standard error is

$$
\operatorname{se}(\widehat\varepsilon)
=\left(\frac{`r round(mean_sqft, 4)`}{`r round(mean_price, 4)`}\right)
`r round(se_b2_house, 4)`
=`r round(se_elasticity, 4)`.
$$

Thus, the 95 percent confidence interval is

$$
`r round(elasticity_hat, 4)`
\pm `r round(tc_elasticity, 4)`\times`r round(se_elasticity, 4)`
= [`r round(elasticity_ci["lower"], 4)`,\;
`r round(elasticity_ci["upper"], 4)`].
$$

#### 3. Interpretation

A 100-square-foot increase in floor area is associated with approximately
13,403 dollars higher expected sale price. At the sample means, one percent
greater floor area is associated with approximately 1.46 percent higher
expected price. The 95 percent confidence interval for the elasticity is
approximately 1.37 to 1.56.

### 3.22(b): Is the elasticity equal to one?

We test at the 1 percent level

$$
H_0:\varepsilon=1
\qquad\text{against}\qquad
H_1:\varepsilon\neq1.
$$

#### 1. Calculate the test in R

```{r}
#| label: solution-322b-en

elasticity_null <- 1
alpha_elasticity <- 0.01

t_elasticity <-
  (elasticity_hat - elasticity_null) / se_elasticity

tc_elasticity_test <-
  qt(1 - alpha_elasticity / 2, df = df_house)

p_elasticity <-
  2 * pt(-abs(t_elasticity), df = df_house)

c(
  t_statistic = t_elasticity,
  lower_critical_value = -tc_elasticity_test,
  upper_critical_value = tc_elasticity_test,
  p_value = p_elasticity,
  reject_null = p_elasticity <= alpha_elasticity
)
```

#### 2. Substitute the numbers into the test statistic

$$
t
=\frac{\widehat\varepsilon-1}
{\operatorname{se}(\widehat\varepsilon)}
=\frac{`r round(elasticity_hat, 4)`-1}{`r round(se_elasticity, 4)`}
=`r round(t_elasticity, 4)`.
$$

The critical values are
$\pm t(0.995,498)=\pm`r round(tc_elasticity_test, 4)`$, and

$$
p=2P\left[T(498)\geq |`r round(t_elasticity, 4)`|\right]
=`r format(p_elasticity, scientific = TRUE, digits = 3)`.
$$

#### 3. Decision and interpretation

We reject $H_0$ at the 1 percent level. The data provide very strong evidence
that the elasticity at the sample means is not one. The estimate indicates an
elasticity greater than one.

### 3.22(c): Is the marginal effect greater than 13,000 dollars?

We test at the 5 percent level

$$
H_0:\beta_2\leq13
\qquad\text{against}\qquad
H_1:\beta_2>13.
$$

The null value is 13 because `price` is measured in thousands of dollars. A
one-unit increase in `sqft` corresponds to 100 square feet.

#### 1. Calculate the test in R

```{r}
#| label: solution-322c-en

beta_null_house <- 13
alpha_house <- 0.05

t_house <- (b2_house - beta_null_house) / se_b2_house
tc_house <- qt(1 - alpha_house, df = df_house)
p_house <- pt(t_house, df = df_house, lower.tail = FALSE)

c(
  t_statistic = t_house,
  critical_value = tc_house,
  p_value = p_house,
  reject_null = p_house <= alpha_house
)
```

#### 2. Substitute the numbers into the test statistic

$$
t
=\frac{b_2-13}{\operatorname{se}(b_2)}
=\frac{`r round(b2_house, 4)`-13}{`r round(se_b2_house, 4)`}
=`r round(t_house, 4)`.
$$

The critical value is $t(0.95,498)=`r round(tc_house, 4)`$, and

$$
p=P\left[T(498)\geq `r round(t_house, 4)`\right]
=`r round(p_house, 4)`.
$$

#### 3. Decision and interpretation

We do not reject $H_0$. Although the point estimate is above 13,000 dollars,
the difference is not sufficiently large relative to its standard error to
provide evidence that the population marginal effect exceeds 13,000 dollars.

### 3.22(d): Expected price of a 2,000-square-foot house

#### 1. Manual calculation using the model estimates

```{r}
#| label: solution-322d-en

house_2000 <- data.frame(sqft = 20)

b1_house <- coef(house_model)["(Intercept)"]
expected_price_manual <- b1_house + 20 * b2_house

sigma_hat_house <- summary(house_model)$sigma
sum_sqft_squared <- sum((collegetown$sqft - mean_sqft)^2)

se_expected_price_2000 <- sigma_hat_house * sqrt(
  1 / nrow(collegetown) +
    (20 - mean_sqft)^2 / sum_sqft_squared
)

tc_price <- qt(0.975, df = df_house)

price_ci_manual <- c(
  lower = expected_price_manual - tc_price * se_expected_price_2000,
  estimate = expected_price_manual,
  upper = expected_price_manual + tc_price * se_expected_price_2000
)

price_ci_manual

predict(
  house_model,
  newdata = house_2000,
  interval = "confidence",
  level = 0.95
)
```

The fitted expected price is

$$
`r round(b1_house, 4)`+20\times`r round(b2_house, 4)`
=`r round(expected_price_manual, 4)`
$$

thousand dollars. Its standard error is

$$
\hat\sigma
\sqrt{\frac{1}{N}+
\frac{(20-\overline{SQFT})^2}
{\sum_i(SQFT_i-\overline{SQFT})^2}}
=`r round(se_expected_price_2000, 4)`.
$$

The 95 percent interval is

$$
`r round(expected_price_manual, 4)`
\pm `r round(tc_price, 4)`\times`r round(se_expected_price_2000, 4)`
= [`r round(price_ci_manual["lower"], 4)`,\;
`r round(price_ci_manual["upper"], 4)`].
$$

The estimated expected price is approximately 152,635 dollars, with a 95
percent confidence interval from approximately 141,550 to 163,721 dollars.
This is an interval for the expected price, not for one individual house.

### 3.22(e): Compare with observed houses

```{r}
#| label: solution-322e-en

houses_observed_2000 <- subset(collegetown, sqft == 20)

houses_observed_2000[, c("price", "sqft")]

c(
  mean_price = mean(houses_observed_2000$price),
  number_of_houses = nrow(houses_observed_2000)
)
```

For the three observed houses,

$$
\overline{PRICE}_{SQFT=20}
=\frac{138+169+183}{3}
=`r round(mean(houses_observed_2000$price), 4)`
$$

thousand dollars. This mean lies inside the confidence interval from part (d),
although near its upper endpoint. The comparison should be interpreted with
caution because the observed mean is based on only three houses.

## Suggested timing

| Activity | Time |
|---|---:|
| Review of key concepts | 5 min |
| Critical-value warm-up | 8 min |
| Worked `food` example | 22 min |
| Transport exercise (a)–(b) | 10 min |
| Transport exercise (c)–(d) | 20 min |
| Class review and comparison | 17 min |
| Left-tailed exit exercise | 5 min |
| Optional Exercise 3.22 | If time permits or homework |
