---
title: "Seminar: Confidence Intervals and Hypothesis Testing"
subtitle: "Chapter 3 – student version"
lang: en
format:
  html:
    toc: true
    code-copy: true
    embed-resources: true
execute:
  warning: true
  message: false
---

## Learning objectives

After the seminar, you should be able to

- construct and interpret a confidence interval for a regression coefficient,
- formulate a null and an alternative hypothesis,
- carry out a one-sided or two-sided $t$-test,
- calculate and interpret a p-value, and
- state the conclusion in its economic context.

The general workflow is

$$
\text{economic question}
\longrightarrow H_0\text{ and }H_1
\longrightarrow t\text{-value}
\longrightarrow \text{p-value}
\longrightarrow \text{conclusion}.
$$

```{r}
#| label: setup
#| include: false

options(scipen = 10)
options(digits = 4)


```

## Five-minute review

### Confidence intervals

Let $b_k$ estimate $\beta_k$, with estimated standard error $\operatorname{se}(b_k)$. A $100(1-\alpha)$ percent confidence interval is

$$
b_k \pm t_{(1-\alpha/2,\,df)}\operatorname{se}(b_k).
$$

The textbook uses $t_{(p,df)}$ for the $p$th percentile of a
$t$-distribution with $df$ degrees of freedom. We use $t_c$ as a short
name for the critical value. For a two-sided interval,

$$
t_c=t_{(1-\alpha/2,df)}.
$$

The method produces intervals that contain the true parameter in $100(1-\alpha)$ percent of repeated samples. It is not correct to say that there is a 95 percent probability that a fixed parameter lies in one already calculated interval.

### Hypothesis tests

For $H_0:\beta_k=c$, the test statistic is

$$
t=\frac{b_k-c}{\operatorname{se}(b_k)}.
$$

If $H_0$ is true and the model assumptions hold, it follows a $t$-distribution with $N-2$ degrees of freedom.

| Alternative         | Test         | Extreme outcomes are in |
|---------------------|--------------|-------------------------|
| $H_1:\beta_k>c$     | Right-tailed | Right tail              |
| $H_1:\beta_k<c$     | Left-tailed  | Left tail               |
| $H_1:\beta_k\neq c$ | Two-sided    | Both tails              |

The direction must be determined by the economic question before examining the estimate.

### P-values

The p-value is the probability, **assuming that the null hypothesis is true**, of obtaining a test statistic at least as extreme as the observed value in the direction specified by $H_1$.

Reject $H_0$ if $p\leq\alpha$. If we do not reject $H_0$, the data do not provide sufficient evidence against it. This does not prove that $H_0$ is true.

## Warm-up: critical values in R

Suppose that $N=32$. In a simple linear regression with an intercept and one explanatory variable, the degrees of freedom are

$$
df=N-2=30.
$$

Use R to find

1.  the critical value for a 95 percent confidence interval,
2.  the critical value for a right-tailed test with $\alpha=0.05$,
3.  the critical value for a left-tailed test with $\alpha=0.05$, and
4.  both critical values for a two-sided test with $\alpha=0.05$.

```{r}
#| label: critical-values-warm-up

N <- 32
df <- N - 2
alpha <- 0.05

critical_values <- c(
  confidence_interval_95 = qt(1 - alpha / 2, df = df),
  right_tailed_test = qt(1 - alpha, df = df),
  left_tailed_test = qt(alpha, df = df),
  two_sided_lower = qt(alpha / 2, df = df),
  two_sided_upper = qt(1 - alpha / 2, df = df)
)

critical_values
```

The location of the alternative hypothesis determines the tail or tails and, therefore, which percentile is required:

$$
H_1
\longrightarrow
\text{tail or tails}
\longrightarrow
\text{critical value}.
$$

## Worked example: food expenditure and income

Consider

$$
FOOD\_EXP_i=\beta_1+\beta_2INCOME_i+e_i.
$$

`food_exp` is weekly food expenditure in dollars. `income` is weekly household income measured in hundreds of dollars. Thus, $\beta_2$ measures the change in expected weekly food expenditure associated with a 100-dollar increase in weekly income.

### Estimate the model

```{r}
#| label: load-and-estimate-food

load(url("http://www.principlesofeconometrics.com/poe5/data/rdata/food.rdata"))

food_model <- lm(food_exp ~ income, data = food)
summary(food_model)

b2_food <- coef(food_model)["income"]
se_b2_food <- coef(summary(food_model))["income", "Std. Error"]
df_food <- df.residual(food_model)

c(
  estimate = b2_food,
  standard_error = se_b2_food,
  degrees_of_freedom = df_food
)
```

### A 95 percent confidence interval

#### 1. Calculate it in R

```{r}
#| label: food-confidence-interval-en

alpha_ci <- 0.05
t_c_ci <- qt(1 - alpha_ci / 2, df = df_food)

food_ci <- c(
  lower = b2_food - t_c_ci * se_b2_food,
  estimate = b2_food,
  upper = b2_food + t_c_ci * se_b2_food
)

food_ci
confint(food_model, "income", level = 0.95)
```

#### 2. Connect the code to the formula

$$
\begin{aligned}
b_2\pm t_c\operatorname{se}(b_2)
&=`r round(b2_food, 4)`
\pm `r round(t_c_ci, 4)`\times`r round(se_b2_food, 4)`\\
&=[`r round(food_ci[1], 4)`,\;
`r round(food_ci[3], 4)`].
\end{aligned}
$$

#### 3. Interpret the result

A 100-dollar increase in weekly income is associated with an estimated `r round(b2_food, 2)`-dollar increase in expected weekly food expenditure. The 95 percent confidence interval ranges from approximately `r round(food_ci[1], 2)` to `r round(food_ci[3], 2)` dollars. This is an estimated association; the regression alone does not establish a causal effect.

### Test 1: Is the slope different from zero?

We test

$$
H_0:\beta_2=0
\qquad\text{against}\qquad
H_1:\beta_2\neq0.
$$

This is a two-sided test because both positive and negative departures from zero count as evidence against $H_0$.

```{r}
#| label: food-two-sided-test-en

null_1 <- 0
alpha_1 <- 0.05

t_food_1 <- (b2_food - null_1) / se_b2_food
t_c_food_1 <- qt(1 - alpha_1 / 2, df = df_food)
p_food_1 <- 2 * pt(-abs(t_food_1), df = df_food)

c(
  t_statistic = t_food_1,
  lower_critical_value = -t_c_food_1,
  upper_critical_value = t_c_food_1,
  p_value = p_food_1
)
```

The calculation is

$$
t=\frac{b_2-0}{\operatorname{se}(b_2)}
=\frac{`r round(b2_food, 4)`-0}{`r round(se_b2_food, 4)`}
=`r round(t_food_1, 4)`.
$$

The two-sided p-value is

$$
p=2P\left[T(38)\geq |`r round(t_food_1, 4)`|\right]
=`r format(p_food_1, scientific = TRUE, digits = 3)`.
$$

We reject $H_0$ at the 5 percent level. The data provide statistical evidence of an association between income and expected food expenditure.

### Test 2: Is the increase greater than 5.50 dollars?

Suppose profitability requires expected food expenditure to rise by more than 5.50 dollars when weekly income rises by 100 dollars. We test

$$
H_0:\beta_2\leq5.5
\qquad\text{against}\qquad
H_1:\beta_2>5.5.
$$

```{r}
#| label: food-right-tailed-test-en

null_2 <- 5.5
alpha_2 <- 0.05

t_food_2 <- (b2_food - null_2) / se_b2_food
t_c_food_2 <- qt(1 - alpha_2, df = df_food)
p_food_2 <- pt(t_food_2, df = df_food, lower.tail = FALSE)

c(
  t_statistic = t_food_2,
  critical_value = t_c_food_2,
  p_value = p_food_2,
  reject_null = p_food_2 <= alpha_2
)
```

The calculation is

$$
t=\frac{b_2-5.5}{\operatorname{se}(b_2)}
=\frac{`r round(b2_food, 4)`-5.5}{`r round(se_b2_food, 4)`}
=`r round(t_food_2, 4)`.
$$

Because the test is right-tailed,

$$
p=P\left[T(38)\geq `r round(t_food_2, 4)`\right]
=`r round(p_food_2, 4)`.
$$

We reject $H_0$ at the 5 percent level, but not at the 1 percent level. The data provide evidence at the 5 percent level that the increase exceeds 5.50 dollars.

For comparison, the two-sided p-value for the same null value is

```{r}
#| label: compare-food-p-values-en

p_food_2_two_sided <- 2 * pt(-abs(t_food_2), df = df_food)

c(
  right_tailed_p_value = p_food_2,
  two_sided_p_value = p_food_2_two_sided
)
```

The test statistic does not change. The p-value changes because the two-sided test includes extreme outcomes in both tails.

::: callout-important
## Before starting the student exercise

For every test, write down $H_0$ and $H_1$ before calculating the p-value. Finish with one statistical conclusion and one interpretation in the economic context.
:::

## Student exercise: transport expenditure and income

This is a constructed teaching dataset containing 32 households. Each
row represents one household:

- `income` is weekly household income measured in hundreds of dollars.
- `transport` is weekly household transport expenditure in dollars.

Transport expenditure may include public transport, fuel, and taxi
fares. The data were constructed for practising statistical inference;
they do not come from an actual household survey.

We examine whether households with higher income also have higher
expected transport expenditure using the model

$
TRANSPORT_i=\beta_1+\beta_2 INCOME_i+e_i.
$

Use both the mathematical formulas and R. For every interval and test,
finish with

1. a statistical decision, and
2. an interpretation in the context of transport expenditure.

::: callout-tip
## Suggested workflow

Write the hypotheses and identify the tail or tails before calculating the test statistic. Then calculate the critical value and p-value. Do not choose the direction of a test after seeing the estimate.
:::

### (a) Data, interpretation, and degrees of freedom

The code below creates the observations as a `data.frame` and estimates
the model. Consequently, every object used later in the exercise is
created here.

```{r}
#| label: transport-setup-student

transport_data <- data.frame(
  income = rep(4:19, each = 2),
  transport = c(
    51.0, 118.4, 170.0, 117.1, 99.1, 180.4, 40.6, 99.5,
    88.9, 188.1, 156.7, 144.8, 208.1, 116.1, 212.5, 203.9,
    197.4, 137.9, 131.9, 172.8, 156.9, 101.5, 131.5, 211.4,
    211.9, 189.4, 189.0, 178.0, 206.9, 176.0, 304.5, 170.5
  )
)

transport_model <- lm(transport ~ income, data = transport_data)
summary(transport_model)

plot(
  transport ~ income,
  data = transport_data,
  xlab = "Weekly income (hundreds of dollars)",
  ylab = "Weekly transport expenditure (dollars)",
  pch = 16
)
abline(transport_model, col = "steelblue", lwd = 2)

b1 <- coef(transport_model)["(Intercept)"]
b2 <- coef(transport_model)["income"]
se_b1 <- coef(summary(transport_model))["(Intercept)", "Std. Error"]
se_b2 <- coef(summary(transport_model))["income", "Std. Error"]
df <- df.residual(transport_model)

c(
  b1 = b1,
  se_b1 = se_b1,
  b2 = b2,
  se_b2 = se_b2,
  df = df
)
```

1. How many observations and variables are in `transport_data`?
2. Use the figure to assess whether the association appears positive or
   negative.
3. Interpret $b_1$ and $b_2$ using the correct units.
4. State the degrees of freedom.

Remember that a one-unit increase in `income` means a 100-dollar
increase in weekly income.

### (b) A 95 percent confidence interval

Construct and interpret a 95 percent confidence interval for $\beta_2$.

Use

$$
b_2\pm t_{(1-\alpha/2,N-2)}\operatorname{se}(b_2).
$$

The code is fully set up and uses only objects created from
`transport_model` in part (a):

```{r}
#| label: transport-confidence-interval-student

alpha <- 0.05
t_c <- qt(1 - alpha / 2, df = df)

transport_ci <- c(
  lower = b2 - t_c * se_b2,
  estimate = b2,
  upper = b2 + t_c * se_b2
)

transport_ci
```

Also show the calculation by substituting the estimated values into

$$
b_2\pm t_{(0.975,30)}\operatorname{se}(b_2).
$$

**Your interpretation:** What does the interval say about the change in
expected weekly transport expenditure associated with a 100-dollar
increase in weekly income?

### (c) A right-tailed economic hypothesis

A transport company argues that a project is profitable only if a 100-dollar increase in weekly income raises expected weekly transport expenditure by more than 4 dollars.

Test at the 5 percent level

$$
H_0:\beta_2\leq4
\qquad\text{against}\qquad
H_1:\beta_2>4.
$$

The test statistic is

$$
t=\frac{b_2-4}{\operatorname{se}(b_2)}
=\frac{`r round(b2, 4)`-4}{`r round(se_b2, 4)`}.
$$

Complete the code:

```{r}
#| label: transport-right-test-student
#| eval: false

beta_null <- 4
alpha <- 0.05

t_obs <- (b2 - beta_null) / se_b2
t_c <- qt(___, df = df)
p_value <- pt(t_obs, df = df, lower.tail = ___)

c(
  t_statistic = t_obs,
  critical_value = t_c,
  p_value = p_value,
  reject_null = p_value <= alpha
)
```

Answer the following:

1.  Why is this a right-tailed test?
2.  What is the rejection region?
3.  Do you reject or not reject $H_0$ at the 5 percent level?
4.  What is the economic conclusion?
5.  Would the decision change at the 10 percent level? Use the same p-value to explain.

### (d) A two-sided hypothesis

Test at the 5 percent level

$$
H_0:\beta_2=8
\qquad\text{against}\qquad
H_1:\beta_2\neq8.
$$

Start with

$$
t=\frac{b_2-8}{\operatorname{se}(b_2)}
=\frac{`r round(b2, 4)`-8}{`r round(se_b2, 4)`}.
$$

Complete the code:

```{r}
#| label: transport-two-sided-test-student
#| eval: false

beta_null <- 8
alpha <- 0.05

t_obs <- (b2 - beta_null) / se_b2
t_c <- qt(___, df = df)
p_value <- 2 * pt(___, df = df)

c(
  t_statistic = t_obs,
  lower_critical_value = ___,
  upper_critical_value = ___,
  p_value = p_value,
  reject_null = ___
)
```

Report the test statistic, critical values, p-value, decision, and economic conclusion.

### (e) Compare the tests

Answer briefly:

1.  Why do parts (c) and (d) have different p-values even though they use the same estimated slope and standard error?
2.  Why can a hypothesis be rejected at the 10 percent level but not at the 5 percent level?
3.  Which hypothesis test would `summary(model)` report automatically for a slope coefficient?
4.  Why would that automatic p-value not answer either part (c) or part (d)?

## Short exit exercise: a left-tailed test

Suppose a demand model gives

$$
b_2=-0.90,\qquad
\operatorname{se}(b_2)=0.35,\qquad
N=25.
$$

At the 5 percent level, test

$$
H_0:\beta_2\geq0
\qquad\text{against}\qquad
H_1:\beta_2<0.
$$

```{r}
#| label: left-tailed-exit-student
#| eval: false

b2 <- -0.90
se_b2 <- 0.35
df <- 25 - 2
alpha <- 0.05

t_obs <- (b2 - 0) / se_b2
t_c <- qt(___, df = df)
p_value <- pt(___, df = df)

c(
  t_statistic = t_obs,
  critical_value = t_c,
  p_value = p_value,
  reject_null = ___
)
```

State the decision and interpret it in terms of price and expected demand.

## Extra Exercise 3.22: house prices in Collegetown

Use this exercise if you finish the transport exercise early. The dataset contains 500 sales of single-family houses in Baton Rouge between 2009 and 2013.

- `price` is the sale price in thousands of dollars.
- `sqft` is floor area in hundreds of square feet.

We use

$$
PRICE_i=\beta_1+\beta_2SQFT_i+e_i.
$$

### Preparation

The data are downloaded directly from the textbook website:

```{r}
#| label: load-collegetown-student

load(url(
  "http://www.principlesofeconometrics.com/poe5/data/rdata/collegetown.rdata"
))

head(collegetown)

plot(
  x = collegetown$sqft,
  y = collegetown$price,
  xlab = "Floor area (hundreds of square feet)",
  ylab = "Sale price (thousands of dollars)",
  main = "House price and floor area",
  pch = 16,
  col = rgb(0, 0, 0, 0.35)
)
```

### 3.22(a): Estimate the model and elasticity

1.  Estimate the model using `lm()`.
2.  Interpret the slope using the correct units.
3.  Calculate the price elasticity with respect to floor area at the sample means:

$$
\widehat\varepsilon
=b_2\frac{\overline{SQFT}}{\overline{PRICE}}.
$$

Treating the sample means as fixed numbers, use

$$
\operatorname{se}(\widehat\varepsilon)
=\frac{\overline{SQFT}}{\overline{PRICE}}
\operatorname{se}(b_2).
$$

4.  Construct a 95 percent confidence interval for the elasticity.
5.  Interpret the point estimate and interval.

The model specification and extraction commands are partly completed:

```{r}
#| label: exercise-322a-student
#| eval: false

house_model <- lm(price ~ sqft, data = collegetown)
summary(house_model)

b2_house <- coef(house_model)["sqft"]
se_b2_house <- coef(summary(house_model))["sqft", "Std. Error"]
df_house <- df.residual(house_model)

mean_sqft <- mean(collegetown$sqft)
mean_price <- mean(collegetown$price)

elasticity_scale <- ___
elasticity_hat <- ___
se_elasticity <- ___

t_c_elasticity <- qt(___, df = df_house)
elasticity_ci <- c(
  lower = ___,
  estimate = elasticity_hat,
  upper = ___
)

elasticity_ci
```

### 3.22(b): Is the elasticity equal to one?

At the 1 percent level, test

$$
H_0:\varepsilon=1
\qquad\text{against}\qquad
H_1:\varepsilon\neq1.
$$

```{r}
#| label: exercise-322b-student
#| eval: false

elasticity_null <- 1
alpha_elasticity <- 0.01

t_elasticity <-
  (elasticity_hat - elasticity_null) / se_elasticity

t_c_elasticity_test <- qt(___, df = df_house)
p_elasticity <- 2 * pt(___, df = df_house)

c(
  t_statistic = t_elasticity,
  lower_critical_value = ___,
  upper_critical_value = ___,
  p_value = p_elasticity,
  reject_null = ___
)
```

Explain why the test is two-sided and report the statistical and economic conclusions.

### 3.22(c): Is the marginal effect greater than 13,000 dollars?

At the 5 percent level, test

$$
H_0:\beta_2\leq13
\qquad\text{against}\qquad
H_1:\beta_2>13.
$$

The null value is 13 because `price` is measured in thousands of dollars and one unit of `sqft` represents 100 square feet.

```{r}
#| label: exercise-322c-student
#| eval: false

beta_null_house <- 13
alpha_house <- 0.05

t_house <- (b2_house - beta_null_house) / se_b2_house
t_c_house <- qt(___, df = df_house)
p_house <- pt(___, df = df_house, lower.tail = ___)

c(
  t_statistic = t_house,
  critical_value = t_c_house,
  p_value = p_house,
  reject_null = ___
)
```

### 3.22(d): Expected price of a 2,000-square-foot house

A 2,000-square-foot house has `sqft = 20`.

1.  Estimate its expected price.
2.  Construct a 95 percent confidence interval for the expected price.
3.  Convert the results from thousands of dollars to dollars in your interpretation.

```{r}
#| label: exercise-322d-student
#| eval: false

house_2000 <- data.frame(sqft = 20)

expected_price_2000 <- predict(
  house_model,
  newdata = house_2000,
  interval = ___,
  level = 0.95
)

expected_price_2000
```

Use `interval = "confidence"` because the question concerns an expected price rather than the price of one new individual house.

### 3.22(e): Compare with observed houses

```{r}
#| label: exercise-322e-student
#| eval: false

houses_observed_2000 <- subset(collegetown, sqft == 20)

houses_observed_2000[, c("price", "sqft")]
mean(houses_observed_2000$price)
nrow(houses_observed_2000)
```

Compare the observed mean with the estimate and interval from part (d). Why should a mean based on very few houses be interpreted cautiously?

## Extra Exercise 3.23: a quadratic house-price model

This exercise continues with the `collegetown` data from Exercise 3.22.
We use the quadratic model

$$
PRICE_i=\alpha_1+\alpha_2 SQFT_i^2+e_i.
$$

Because `price` is measured in thousands of dollars and `sqft` in
hundreds of square feet, the marginal effect at $SQFT=x_0$ is

$$
\frac{dE(PRICE\mid SQFT)}{dSQFT}\bigg|_{x_0}
=2\alpha_2x_0.
$$

The estimated marginal effect and its standard error are therefore

$$
\widehat{ME}(x_0)=2a_2x_0,
\qquad
\operatorname{se}\!\left[\widehat{ME}(x_0)\right]
=|2x_0|\operatorname{se}(a_2).
$$

### Prepare the model

```{r}
#| label: exercise-323-setup-en
#| eval: false

quadratic_model <- lm(price ~ I(sqft^2), data = collegetown)
summary(quadratic_model)

a1 <- coef(quadratic_model)["(Intercept)"]
a2 <- coef(quadratic_model)["I(sqft^2)"]
se_a2 <- coef(summary(quadratic_model))["I(sqft^2)", "Std. Error"]
df_quadratic <- df.residual(quadratic_model)
```

### 3.23(a): Marginal effect at 2,000 square feet

A 2,000-square-foot house has `sqft = 20`. At the 5 percent level, test

$$
H_0:2\alpha_2(20)\leq13
\qquad\text{against}\qquad
H_1:2\alpha_2(20)>13.
$$

```{r}
#| label: exercise-323a-student-en
#| eval: false

x0 <- 20
marginal_effect_20 <- 2 * a2 * x0
se_marginal_20 <- abs(2 * x0) * se_a2

t_20 <- (marginal_effect_20 - 13) / se_marginal_20
t_c_20 <- qt(___, df = df_quadratic)
p_20 <- pt(___, df = df_quadratic, lower.tail = ___)

c(
  marginal_effect = marginal_effect_20,
  standard_error = se_marginal_20,
  t_statistic = t_20,
  critical_value = t_c_20,
  p_value = p_20,
  reject_null = ___
)
```

Report the test statistic, rejection region, p-value, and economic
conclusion.

### 3.23(b): Marginal effect at 4,000 square feet

Repeat the test for a 4,000-square-foot house, so that `sqft = 40`.
Explain why both the marginal effect and its standard error change when
$x_0$ changes.

```{r}
#| label: exercise-323b-student-en
#| eval: false

x0 <- 40
marginal_effect_40 <- ___
se_marginal_40 <- ___

t_40 <- ___
t_c_40 <- ___
p_40 <- ___
```

### 3.23(c): Expected price at 2,000 square feet

Estimate the expected price and construct a 95 percent confidence
interval.

```{r}
#| label: exercise-323c-student-en
#| eval: false

house_2000_quadratic <- data.frame(sqft = 20)

expected_price_quadratic <- predict(
  quadratic_model,
  newdata = house_2000_quadratic,
  interval = "confidence",
  level = 0.95
)

expected_price_quadratic
```

Interpret the result in dollars rather than thousands of dollars.

### 3.23(d): Compare with observed houses

Use the houses with `sqft == 20` found in Exercise 3.22(e). Is their
mean sale price compatible with the interval from 3.23(c)? Explain.

## Final checklist

Before finishing, check that you have

- used the correct variable units,
- stated both $H_0$ and $H_1$,
- selected the correct tail or tails,
- compared the p-value with the stated significance level,
- written “reject” or “do not reject,” and
- interpreted the result in its economic context.
